### 2023-05-01 chapter0/chapter0.tex
Added subsections to discussions

\section{Aims}

Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between the sequence and function remains poorly understood. In contrast, predictions of structure and function are readily available for many folded domains. Because they make specific contacts between residues, the sequences of folded domains are generally conserved or evolve slowly. Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}. Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone was an active area of research for decades. Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences. Even for a moderately sized protein of 100 amino acid residues, there are $20^{100} \approx 1.3 \times 10^{130}$ possible sequences. Only in the past few years have researchers in many senses solved this problem by using advanced techniques from machine learning to leverage the information encoded in nearly two hundred thousand experimentally determined structures~\cite{Jumper2021}.

>>>>

\section{Aims}
Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between the sequence and function remains poorly understood. In contrast, predictions of structure and function are readily available for many folded domains. Because they make specific contacts between residues, the sequences of folded domains are generally conserved or evolve slowly. Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}. Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone was an active area of research for decades. Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences. Even for a moderately sized protein of 100 amino acid residues, there are $20^{100} \approx 1.3 \times 10^{130}$ possible sequences. Only in the past few years have researchers in many senses solved this problem by using advanced techniques from machine learning to leverage the information encoded in nearly two hundred thousand experimentally determined structures~\cite{Jumper2021}.

### 2023-05-01 chapter1/chapter1.tex
Added subsections to discussions

\section{Discussion}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus. To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data. Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well. While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated. However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work. Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.

@@

While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

@@

Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid~\cite{Kaduk2017}, OMA standalone~\cite{Altenhoff2019}, OrthoFinder~\cite{Emms2019}, OrthoInspector~\cite{Linard2014}, Orthologer~\cite{Zdobnov2020}), we have chosen not to package the code into an end-to-end pipeline. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

>>>>

\section{Discussion}
\subsection{Developments in genome assembly and annotation challenge existing methods of orthology inference}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus. To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data. Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well. While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated. However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work. Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.

@@

\subsection{Comprehensive bioinformatic analyses of proteins will depend on splice-sensitive alignments}
While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller~\cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

@@

\subsection{Accessible computational tools will facilitate future comparative studies}
Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid~\cite{Kaduk2017}, OMA standalone~\cite{Altenhoff2019}, OrthoFinder~\cite{Emms2019}, OrthoInspector~\cite{Linard2014}, Orthologer~\cite{Zdobnov2020}), we have chosen not to package the code into an end-to-end pipeline. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

### 2023-05-01 chapter2/chapter2.tex
Added subsections to discussions

\subsection{A workflow for future evolutionary analyses of IDRs}
As discussed by Zarin \textit{et al.}, the interpretation of evolutionary signatures is complicated by several methodological limitations~\cite{Zarin2019}. For example, because IDRs are identified as contiguous segments of high predicted disorder, their boundaries are defined by adjacent structural elements. This approach can therefore split an IDR that is a single evolutionary or functional unit if it contains a semi-disordered module that scores below the threshold. Conversely, it can also merge two distinct IDRs if they are not separated by at least one folded domain. Another challenge is the significant overlap in the definitions of many features induces strong correlations that preclude straightforward quantitative manipulations or interpretations of an IDR's evolutionary signature. The original authors have since addressed this in subsequent studies by applying machine learning methods to perform feature selection or learn features directly from alignments of IDRs~\cite{Zarin2021, Lu2022}. However, integrating these methods into a unified phylogenetic comparative framework will require further effort.

>>>>

\subsection{Future evolutionary analyses of IDRs will require a multimodal approach}
As discussed by Zarin \textit{et al.}, the interpretation of evolutionary signatures is complicated by several methodological limitations~\cite{Zarin2019}. For example, because IDRs are identified as contiguous segments of high predicted disorder, their boundaries are defined by adjacent structural elements. This approach can therefore split an IDR that is a single evolutionary or functional unit if it contains a semi-disordered module that scores below the threshold. Conversely, it can also merge two distinct IDRs if they are not separated by at least one folded domain. Another challenge is the significant overlap in the definitions of many features induces strong correlations that preclude straightforward quantitative manipulations or interpretations of an IDR's evolutionary signature. The original authors have since addressed this in subsequent studies by applying machine learning methods to perform feature selection or learn features directly from alignments of IDRs~\cite{Zarin2021, Lu2022}. However, integrating these methods into a unified phylogenetic comparative framework will require further effort.
